---
title: "ggplot2"
subtitle: "Quick Introduction"
author: "Berk Orbay"
institute: "MEF BDA 503"
format: 
    revealjs:
        smaller: true
---

## What is dplyr?

`ggplot2` is an R package specialized on data visualization. It is both a companion to the `dplyr` data manipulation package and a whole library by itself. Within a well defined framework, you can

+ easily draw a large number plot types with the same methodology
+ draw sophisticated plots
+ multiple well adjusted plots (e.g, 2x2)
+ do advanced theming and custom theming
+ parametric plots

 
\ 

Official page: <https://ggplot2.tidyverse.org/>

## Why ggplot2?

`ggplot2` has the same UX in mind with `dplyr` and is being developed by the same person (Hadley Wickham) and their team.

+ Based on Grammar of Graphics (hence the "gg").
+ Start with a "canvas" (`ggplot()`) and apply data (`aes(x,y,color)`) and layers (`geom_*`) on it
+ Add labels, annotations, titles, themes easily.

# Fundamentals

**Important** Full notes are on [Book of EDA](https://berkorbay.github.io/book-of-eda/ggplot2.html)

\

```{r}
#| eval: true
#| echo: true

##  If you have never installed ggplot2 before
##  install.packages("tidyverse")

##  First load the library
##  We will also use dplyr therefore tidyverse is better
library(tidyverse)

## We are going to use starwars datasets to show examples
starwars
```

Let's recall `starwars` tibble from `dplyr`.

## Scatterplot `geom_point`

`select` simply displays columns

```{r}
#| eval: true
#| echo: true

ggplot(starwars %>% select(name, height, mass) %>% filter(mass < 1000) %>% filter(complete.cases(.)), aes(x = height, y = mass)) +
    geom_point()
```

We use `+` instead of the pipe operator (`%>%`) to connect layers.

## Scatterplot (cont.) `geom_point`

Let's add some categorization with `color`.

```{r}
#| eval: true
#| echo: true

ggplot(starwars %>% select(name, species, height, mass) %>% filter(mass < 1000) %>% filter(complete.cases(.)), aes(x = height, y = mass)) +
    geom_point(aes(color = species))
```

## Scatterplot (cont.) `geom_point`

Let's also add some labels and themes.

```{r}
#| eval: true
#| echo: true

ggplot(starwars %>% select(name, species, height, mass) %>% filter(mass < 1000) %>% filter(complete.cases(.)), aes(x = height, y = mass)) +
    geom_point(aes(color = species)) +
    labs(title = "Star Wars Characters by Height and Mass", subtitle = "Colors by species", x = "Height (in meters)", y = "Mass (in KG)", color = "Species") +
    theme_minimal() +
    theme(legend.position = "bottom")
```


## Advanced `ggplot2`

There is always more to learn. Here are some examples.

+ `facet_grid` for multiple plots ([see](https://ggplot2.tidyverse.org/reference/facet_grid.html)).
+ `scale_*` functions for fine tuning labels, axes etc.
+ custom themes

See extra resources section on Book of EDA.

# Thanks!

Course webpage [https://mef-bda503.github.io/](https://mef-bda503.github.io/).